---
title: "Minimal Set of binnmu Packages"
subtitle: "Combining the R and Debian package systems"
author: "Dirk Eddelbuettel"
date: "2017-Jul-16"
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Step 0: Problem Definition

### Upstream Change

R 3.4.0, released in April, included the following paragraph in its NEWS file:

 * Packages which register native routines for .C or .Fortran need
   to be re-installed for this version (unless installed with
   R-devel SVN revision r72375 or later).

This transition has no fallback behavior (as is more common with R changes) 
and requires a rebuild.  Packages build under older R version still _load_ 
and function partially, but will be unable to access any native (_i.e._, compiled)
routines.

### Impact

For the Debian packages, this means that we need to consider the set of packages which

- match `r-cran-*`, `r-bioc-*` and alike
- contain compiled code (as R-only packages have no native routines)
- use at least one `.C()` or `.Fortran()` (but _not_ `.Call()`) call
- have _not yet_ been recompiled with R 3.4.0 or R 3.4.1

This note computes this set and provides the input for a 
[wanna-build](https://release.debian.org/wanna-build.txt) request.

## Step 1: Reverse Dependencies of R

### Fresh Debian unstable session

For this we drop into a clean Docker container running Debian unstable. Later, we will need
the current sources of the [`RcppAPT`](https://github.com/eddelbuettel/rcppapt) package so 
we start from a local git directory:

```sh
server> cd ~/git && docker run --rm -ti -v $(pwd):/mnt debian:unstable
```

### Update Debian

Inside the Docker container, we update the package information and install what
is needed to build [`RcppAPT`](https://github.com/eddelbuettel/rcppapt) for R.  
This includes Rcpp and `libapt-pkg-dev`.  We also install the data.table package
used for aggregating the (R and Debian) package data computed below.

This step takes a short moment, with the exact time dependent on the network 
connection and other factors.

```sh
docker> apt-get update
docker> apt-get -y dist-upgrade

docker> apt-get -y install r-cran-rcpp r-cran-data.table libapt-pkg-dev less 

docker> cd /mnt 
docker> R CMD INSTALL rcppapt/       # assuming we're above rcppapt
```

### Launch R

#### All Candidates

Inside the same Docker session, we now launch R and run (almost all of) the remainder from R.

```{r, eval=FALSE}
> library(RcppAPT)
> library(data.table)
> rd <- reverseDepends("r-base-core")         # 514 x 2
> rd <- rd[!grepl("r-(doc|base)", rd[,1]), ]  # 506 x 2
> rd <- rd[grepl("^r-", rd[,1]), ]            # 480 x 2
> rd <- rd[order(rd[,2]), ]
> setDT(rd)
```

We use `RcppAPT` to compute the reverse depends of the main R package providing the R engine: `r-base-core`.
Among those (currently) 514 packages are both other packages from the upstream source (`r-base*`, `r-doc*`) which
we exclude first as well as other, non-R-package dependencies (such as `rpy2`) which we also exclude.

This leaves 480 candidate packages out of the initial 514.  The version field tells which r-base-core version
was used to build the package---information we need per the setup described above.

```{r, eval=FALSE}
> rd
                package          version
  1:   r-cran-phylobase 3.2.4.20160406-1
  2:       r-cran-zelig          3.0.2-1
  3:        r-cran-urca          3.4.0-1
  4:     r-cran-tseries 3.3.3.20170413-1
  5:     r-cran-tkrplot 3.3.0.20160615-1
 ---                                    
476:      r-cran-bitops          3.0.1-6
477: r-cran-cairodevice          3.4.0-1
478:        r-cran-boot          3.4.0-1
479:     r-cran-acepack          3.4.0-1
480:       r-cran-abind          3.3.1-1
> 
```

Next we need to filter out two versions with unsortable (_i.e._,non-semantic) version numbers,
and apply a logical filter depending on whether the package was built with R version 3.3.3 or 
earlier, indicating a possibe required rebuild.

```{r, eval=FALSE}
> rd[ version=="3.0.0~20130330-1", version := "3.0.0.20130330-1"]
> rd[ version=="3.2.4-revised-1", version := "3.2.4.1-1"]
> rd[version!="", oldVersion := version  <=  package_version("3.3.3-1")]
> rd[ is.na(oldVersion), oldVersion := FALSE]
> rd
                package          version oldVersion
  1:   r-cran-phylobase 3.2.4.20160406-1       TRUE
  2:       r-cran-zelig          3.0.2-1       TRUE
  3:        r-cran-urca          3.4.0-1      FALSE
  4:     r-cran-tseries 3.3.3.20170413-1      FALSE
  5:     r-cran-tkrplot 3.3.0.20160615-1       TRUE
 ---                                               
476:      r-cran-bitops          3.0.1-6       TRUE
477: r-cran-cairodevice          3.4.0-1      FALSE
478:        r-cran-boot          3.4.0-1      FALSE
479:     r-cran-acepack          3.4.0-1      FALSE
480:       r-cran-abind          3.3.1-1       TRUE
> 
```

To cover some corner case, we derive a `skip` field:

```{r, eval=FALSE}
> rd[ version=="", skip:=TRUE ]
> rd[ is.na(skip), skip:=FALSE]
> rd
                package          version oldVersion  skip
  1:   r-cran-phylobase 3.2.4.20160406-1       TRUE FALSE
  2:       r-cran-zelig          3.0.2-1       TRUE FALSE
  3:        r-cran-urca          3.4.0-1      FALSE FALSE
  4:     r-cran-tseries 3.3.3.20170413-1      FALSE FALSE
  5:     r-cran-tkrplot 3.3.0.20160615-1       TRUE FALSE
 ---                                                     
476:      r-cran-bitops          3.0.1-6       TRUE FALSE
477: r-cran-cairodevice          3.4.0-1      FALSE FALSE
478:        r-cran-boot          3.4.0-1      FALSE FALSE
479:     r-cran-acepack          3.4.0-1      FALSE FALSE
480:       r-cran-abind          3.3.1-1       TRUE FALSE
> 
```

#### Compiled Packages


Next, we find the actual dependencies of each of these packages by constructing
a large regular expression which we feed into `RcppAPT::getDepends()`


```{r, eval=FALSE}
> regexp <- paste(paste0("^", rd[skip==FALSE, package], "$"), collapse="|")
> dep <- getDepends(regexp)
> setDT(dep)
> dep
                  srcpkg              deppkg cmpop          version
   1:  r-bioc-hypergraph         r-base-core     2 3.3.1.20161024-1
   2:  r-bioc-hypergraph             r-api-3     0           (null)
   3:  r-bioc-hypergraph        r-bioc-graph     0           (null)
   4:  r-bioc-hypergraph r-bioc-biocgenerics     0           (null)
   5:  r-bioc-hypergraph        r-cran-runit     0           (null)
  ---                                                              
3690: r-cran-viridislite             r-api-3     0           (null)
3691:      r-cran-xtable         r-base-core     2          3.2.5-1
3692:      r-cran-xtable             r-api-3     0           (null)
3693:   r-cran-pkgkitten         r-base-core     2          3.3.2-1
3694:   r-cran-pkgkitten             r-api-3     0           (null)
> 
```

Next we subset to those have `libc6` as a Depends, meaning they are compiled packages.
This excludes all the R packages having only R code.

```{r, eval=FALSE}
> comp <- dep[deppkg=="libc6"]   # about 241
> comp[, isCompiled:=TRUE]
> comp
                 srcpkg deppkg cmpop version isCompiled
  1:  r-bioc-makecdfenv  libc6     2     2.4       TRUE
  2:       r-cran-bio3d  libc6     2    2.14       TRUE
  3:   r-bioc-rsamtools  libc6     2    2.15       TRUE
  4:     r-cran-foreign  libc6     2    2.14       TRUE
  5:    r-bioc-multtest  libc6     2    2.14       TRUE
 ---                                                   
237:     r-cran-nleqslv  libc6     2     2.4       TRUE
238: r-other-amsmercury  libc6     2    2.14       TRUE
239:         r-cran-gnm  libc6     2     2.4       TRUE
240:         r-cran-gsl  libc6     2     2.4       TRUE
241:         r-cran-gss  libc6     2     2.4       TRUE
> 
```

We are now getting closer.  We set keys on the `data.table` objects, and then do 
an inner join:

```{r, eval=FALSE}
> setkey(comp, srcpkg)
> setkey(rd, package)
> all <- rd[comp[, c(1,5)]]   # inner join (by default on columns with keys)
> all[order(version),]
                      package version oldVersion  skip isCompiled
  1:            r-cran-bitops 3.0.1-6       TRUE FALSE       TRUE
  2:               r-cran-mnp 3.0.2-1       TRUE FALSE       TRUE
  3: r-other-mott-happy.hbrem 3.0.2-1       TRUE FALSE       TRUE
  4:             r-cran-amore 3.1.0-1       TRUE FALSE       TRUE
  5:              r-cran-deal 3.1.0-1       TRUE FALSE       TRUE
 ---                                                             
237:           r-cran-statmod 3.4.1-1      FALSE FALSE       TRUE
238:               r-cran-ape 3.4.1-2      FALSE FALSE       TRUE
239:           r-cran-nleqslv 3.4.1-2      FALSE FALSE       TRUE
240:            r-cran-rmysql 3.4.1-2      FALSE FALSE       TRUE
241:                r-mathlib 3.4.1-2      FALSE FALSE       TRUE
> 
```

We have 241 _potential_ rebuilds, down from 514 reverse depends at the outset.

#### Version check

Next, we can concentrate on those having been built with the older versions requiring
a rebuild:


```{r, eval=FALSE}
> all[oldVersion==TRUE,][order(version),]    # 167
                      package version oldVersion  skip isCompiled
  1:            r-cran-bitops 3.0.1-6       TRUE FALSE       TRUE
  2:               r-cran-mnp 3.0.2-1       TRUE FALSE       TRUE
  3: r-other-mott-happy.hbrem 3.0.2-1       TRUE FALSE       TRUE
  4:             r-cran-amore 3.1.0-1       TRUE FALSE       TRUE
  5:              r-cran-deal 3.1.0-1       TRUE FALSE       TRUE
 ---                                                             
163:           r-cran-rcppgsl 3.3.3-1       TRUE FALSE       TRUE
164:             r-cran-rodbc 3.3.3-1       TRUE FALSE       TRUE
165:         r-cran-snowballc 3.3.3-1       TRUE FALSE       TRUE
166:                r-cran-v8 3.3.3-1       TRUE FALSE       TRUE
167:               r-cran-zoo 3.3.3-1       TRUE FALSE       TRUE
> 
```

Now we are down to 167 packages.

```{r, eval=FALSE}
> all[, cran:=grepl("^r-cran", package) ]
> all[, bioc:=grepl("^r-bioc", package) ]
> all[bioc==TRUE & oldVersion==TRUE,]                # 17 BioC
                  package          version oldVersion  skip isCompiled  cran bioc
 1:           r-bioc-affy 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
 2:         r-bioc-affyio 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
 3:        r-bioc-biobase 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
 4:     r-bioc-biovizbase          3.3.2-1       TRUE FALSE       TRUE FALSE TRUE
 5:         r-bioc-deseq2          3.3.2-1       TRUE FALSE       TRUE FALSE TRUE
 6:        r-bioc-dnacopy 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
 7:          r-bioc-edger          3.3.0-2       TRUE FALSE       TRUE FALSE TRUE
 8:     r-bioc-genefilter          3.3.2-1       TRUE FALSE       TRUE FALSE TRUE
 9:          r-bioc-graph 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
10:     r-bioc-hilbertvis 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
11:          r-bioc-limma          3.3.2-1       TRUE FALSE       TRUE FALSE TRUE
12:     r-bioc-makecdfenv 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
13:       r-bioc-multtest 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
14: r-bioc-preprocesscore 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
15:           r-bioc-rbgl          3.3.2-1       TRUE FALSE       TRUE FALSE TRUE
16:    r-bioc-rtracklayer          3.3.2-1       TRUE FALSE       TRUE FALSE TRUE
17:       r-bioc-snpstats 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
> 
```

Among these are 17 BioConductor packages.  This is a superset as we do not know which of these 
use only `.Call()` meaning that no rebuild would be required.   


```{r, eval=FALSE}
> all[bioc!=TRUE & cran!=TRUE & oldVersion==TRUE,]   # 3 other
                    package version oldVersion  skip isCompiled  cran  bioc
1:       r-other-amsmercury 3.3.2-1       TRUE FALSE       TRUE FALSE FALSE
2:          r-other-iwrlars 3.3.2-1       TRUE FALSE       TRUE FALSE FALSE
3: r-other-mott-happy.hbrem 3.0.2-1       TRUE FALSE       TRUE FALSE FALSE
> 
```

There are also three which are neither BioC nor CRAN.

```{r, eval=FALSE}
> cand <- all[ cran==TRUE & oldVersion==TRUE, ]   # 147
> cand
             package version oldVersion  skip isCompiled cran  bioc
  1:     r-cran-ade4 3.3.2-1       TRUE FALSE       TRUE TRUE FALSE
  2: r-cran-adegenet 3.3.1-1       TRUE FALSE       TRUE TRUE FALSE
  3: r-cran-adephylo 3.3.2-1       TRUE FALSE       TRUE TRUE FALSE
  4:   r-cran-amelia 3.2.3-1       TRUE FALSE       TRUE TRUE FALSE
  5:    r-cran-amore 3.1.0-1       TRUE FALSE       TRUE TRUE FALSE
 ---                                                               
143:    r-cran-vegan 3.3.2-1       TRUE FALSE       TRUE TRUE FALSE
144:     r-cran-vgam 3.3.2-1       TRUE FALSE       TRUE TRUE FALSE
145:     r-cran-xml2 3.3.2-1       TRUE FALSE       TRUE TRUE FALSE
146:     r-cran-yaml 3.3.2-1       TRUE FALSE       TRUE TRUE FALSE
147:      r-cran-zoo 3.3.3-1       TRUE FALSE       TRUE TRUE FALSE
> 
```

We have 147 possible NMUs based off CRAN.

Next, we mix this with information from CRAN.

```{r, eval=FALSE}
> db <- tools::CRAN_package_db()   # CRAN pkge info: N rows x 65 cols
> setDT(db)
> db[, package:=paste0("r-cran-", tolower(Package))]
> setkey(db, package)              # key on package field
> foo <- db[ cand ]                # inner join
> foo[, .(package, Package, Version, NeedsCompilation, oldVersion, skip)]
             package  Package Version NeedsCompilation oldVersion  skip
  1:     r-cran-ade4     ade4   1.7-6              yes       TRUE FALSE
  2: r-cran-adegenet adegenet   2.0.1              yes       TRUE FALSE
  3: r-cran-adephylo adephylo  1.1-10              yes       TRUE FALSE
  4:   r-cran-amelia   Amelia   1.7.4              yes       TRUE FALSE
  5:    r-cran-amore    AMORE  0.2-15              yes       TRUE FALSE
 ---                                                                   
143:    r-cran-vegan    vegan   2.4-3              yes       TRUE FALSE
144:     r-cran-vgam     VGAM   1.0-3              yes       TRUE FALSE
145:     r-cran-xml2     xml2   1.1.1              yes       TRUE FALSE
146:     r-cran-yaml     yaml  2.1.14              yes       TRUE FALSE
147:      r-cran-zoo      zoo   1.8-0              yes       TRUE FALSE
> 
```

This is our set of 147 candidate packages with their CRAN name, Debian name and upstream version.

```{r, eval=FALSE}
> saveRDS(foo[, .(package, Package, Version, NeedsCompilation, oldVersion, skip)], file="debpackages.rds")
```

We save this file to be used on another machine.

## Step 2: Grep

On another machine with access to all CRAN package sources (which I happen to have access to), we use the list 
of 147 candidate packages and run a recursive grep for each.  We store the output from `egrep`, called via `system()`, 
directly in the same data structure:

```{r, eval=FALSE}
deb <- readRDS("debpackages.rds")
for (i in 1:nrow(deb)) { 
    deb[i, "dotCorFortran"] <- if (is.na(deb[i, "Package"])) NA 
                               else system(paste0("egrep -r -q \"\\.(C|Fortran)\\(\" ", deb[i, "Package"], "/R/*"))==0 
}
saveRDS(deb, "debpackagesout.rds")
```

## Step 3: Finalize

We read the data back in and subset on those for which the recursive grep found actual uses of
`.C()` or `.Fortran()`.  The list contains 72 packages.

```{r, eval=FALSE}
> deb <- readRDS("debpackagesout.rds")
> setDT(deb)
> deb[ is.na(deb[, dotCorFortran]) | deb[, dotCorFortran]==TRUE, 1:3]   ## 72
                    package           Package  Version
 1:              r-cran-ade4              ade4    1.7-6
 2:          r-cran-adegenet          adegenet    2.0.1
 3:          r-cran-adephylo          adephylo   1.1-10
 4:            r-cran-amelia            Amelia    1.7.4
 5:            r-cran-bayesm            bayesm    3.1-0
 6:            r-cran-bitops            bitops    1.0-6
 7:     r-cran-blockmodeling     blockmodeling    0.1.9
 8:           r-cran-boolnet           BoolNet    2.1.3
 9:             r-cran-brglm             brglm    0.6.1
10:             r-cran-caret             caret   6.0-76
11:           r-cran-catools           caTools   1.17.1
12:            r-cran-cmprsk            cmprsk    2.2-7
13:              r-cran-coin              coin    1.2-0
14:          r-cran-contfrac          contfrac   1.1-11
15:        r-cran-data.table        data.table   1.10.4
16:              r-cran-deal              deal   1.2-37
17:            r-cran-deldir            deldir   0.1-14
18:           r-cran-desolve           deSolve     1.20
19:       r-cran-dosefinding       DoseFinding   0.9-15
20:               r-cran-eco               eco    3.1-7
21:               r-cran-erm               eRm   0.15-7
22:               r-cran-etm               etm    0.6-2
23:               r-cran-evd               evd    2.3-2
24:              r-cran-expm              expm  0.999-2
25:            r-cran-fields            fields      9.0
26:               r-cran-gam               gam   1.14-4
27:           r-cran-genabel           GenABEL    1.8-0
28:            r-cran-glmnet            glmnet   2.0-10
29:           r-cran-goftest           goftest    1.1-1
30:               r-cran-gsl               gsl 1.9-10.3
31:       r-cran-haplo.stats       haplo.stats    1.7.7
32:              r-cran-hdf5                NA       NA
33:            r-cran-hexbin            hexbin   1.27.1
34:            r-cran-igraph            igraph    1.0.1
35:               r-cran-lhs               lhs     0.14
36:         r-cran-logspline         logspline    2.1.9
37:           r-cran-mapproj           mapproj    1.2-5
38:              r-cran-maps              maps    3.2.0
39:          r-cran-maptools          maptools    0.9-2
40:              r-cran-mcmc              mcmc    0.9-5
41:          r-cran-mcmcpack          MCMCpack    1.4-0
42:      r-cran-medadherence                NA       NA
43:          r-cran-mixtools          mixtools    1.1.0
44:           r-cran-mlbench           mlbench    2.1-1
45:               r-cran-mnp               MNP    3.0-2
46:               r-cran-msm               msm    1.6.4
47:             r-cran-ncdf4             ncdf4     1.16
48:              r-cran-nnls              nnls      1.4
49:          r-cran-pbivnorm          pbivnorm    0.6.0
50:          r-cran-phangorn          phangorn    2.2.0
51:         r-cran-phylobase         phylobase    0.8.4
52:           r-cran-polycub           polyCub    0.6.0
53:         r-cran-princurve         princurve   1.1-12
54:              r-cran-pscl              pscl    1.4.9
55:               r-cran-qtl               qtl   1.41-6
56:      r-cran-randomfields      RandomFields   3.1.50
57: r-cran-randomfieldsutils RandomFieldsUtils   0.3.25
58:      r-cran-randomforest      randomForest   4.6-12
59:      r-cran-raschsampler      RaschSampler    0.8-8
60:             r-cran-rcurl             RCurl 1.95-4.8
61:         r-cran-rniftilib                NA       NA
62:                r-cran-sp                sp    1.2-5
63:              r-cran-spam              spam    2.1-1
64:          r-cran-spatstat          spatstat   1.51-0
65:               r-cran-spc               spc    0.5.3
66:             r-cran-spdep             spdep   0.6-13
67:      r-cran-surveillance      surveillance   1.14.0
68:               r-cran-tgp               tgp   2.4-14
69:        r-cran-tikzdevice        tikzDevice   0.10-1
70:         r-cran-treescape                NA       NA
71:             r-cran-vegan             vegan    2.4-3
72:              r-cran-vgam              VGAM    1.0-3
                     package           Package  Version
> 
```

These 72, along with the 17 BioC and 3 'other' packages are our target set.

```{r, eval=FALSE}
> nmu <- deb[ is.na(deb[, dotCorFortran]) | deb[, dotCorFortran]==TRUE, 1]
> nmu <- rbind(nmu,  all[bioc!=TRUE & cran!=TRUE & oldVersion==TRUE, 1]) # 'other'
> nmu <- rbind(nmu,  all[bioc==TRUE & oldVersion==TRUE, 1])              # BioC
>              
> options("width"=80)
> nmu[[1]]
 [1] "r-cran-ade4"              "r-cran-adegenet"         
 [3] "r-cran-adephylo"          "r-cran-amelia"           
 [5] "r-cran-bayesm"            "r-cran-bitops"           
 [7] "r-cran-blockmodeling"     "r-cran-boolnet"          
 [9] "r-cran-brglm"             "r-cran-caret"            
[11] "r-cran-catools"           "r-cran-cmprsk"           
[13] "r-cran-coin"              "r-cran-contfrac"         
[15] "r-cran-data.table"        "r-cran-deal"             
[17] "r-cran-deldir"            "r-cran-desolve"          
[19] "r-cran-dosefinding"       "r-cran-eco"              
[21] "r-cran-erm"               "r-cran-etm"              
[23] "r-cran-evd"               "r-cran-expm"             
[25] "r-cran-fields"            "r-cran-gam"              
[27] "r-cran-genabel"           "r-cran-glmnet"           
[29] "r-cran-goftest"           "r-cran-gsl"              
[31] "r-cran-haplo.stats"       "r-cran-hdf5"             
[33] "r-cran-hexbin"            "r-cran-igraph"           
[35] "r-cran-lhs"               "r-cran-logspline"        
[37] "r-cran-mapproj"           "r-cran-maps"             
[39] "r-cran-maptools"          "r-cran-mcmc"             
[41] "r-cran-mcmcpack"          "r-cran-medadherence"     
[43] "r-cran-mixtools"          "r-cran-mlbench"          
[45] "r-cran-mnp"               "r-cran-msm"              
[47] "r-cran-ncdf4"             "r-cran-nnls"             
[49] "r-cran-pbivnorm"          "r-cran-phangorn"         
[51] "r-cran-phylobase"         "r-cran-polycub"          
[53] "r-cran-princurve"         "r-cran-pscl"             
[55] "r-cran-qtl"               "r-cran-randomfields"     
[57] "r-cran-randomfieldsutils" "r-cran-randomforest"     
[59] "r-cran-raschsampler"      "r-cran-rcurl"            
[61] "r-cran-rniftilib"         "r-cran-sp"               
[63] "r-cran-spam"              "r-cran-spatstat"         
[65] "r-cran-spc"               "r-cran-spdep"            
[67] "r-cran-surveillance"      "r-cran-tgp"              
[69] "r-cran-tikzdevice"        "r-cran-treescape"        
[71] "r-cran-vegan"             "r-cran-vgam"             
[73] "r-other-amsmercury"       "r-other-iwrlars"         
[75] "r-other-mott-happy.hbrem" "r-bioc-affy"             
[77] "r-bioc-affyio"            "r-bioc-biobase"          
[79] "r-bioc-biovizbase"        "r-bioc-deseq2"           
[81] "r-bioc-dnacopy"           "r-bioc-edger"            
[83] "r-bioc-genefilter"        "r-bioc-graph"            
[85] "r-bioc-hilbertvis"        "r-bioc-limma"            
[87] "r-bioc-makecdfenv"        "r-bioc-multtest"         
[89] "r-bioc-preprocesscore"    "r-bioc-rbgl"             
[91] "r-bioc-rtracklayer"       "r-bioc-snpstats"  
> 
```

We need to retrieve the version number in Debian unstable of these packages by once agaim
relying of a function from [`RcppAPT`](https://github.com/eddelbuettel/rcppapt) 

```{r, eval=FALSE}
> regexp <- paste(paste0("^", nmu[[1]], "$"), collapse="|")
> res <- getPackages(regexp)
> res
> res
                    Package         Version
1         r-bioc-makecdfenv        1.50.0-1
2           r-bioc-multtest        2.30.0-1
3              r-bioc-edger   3.14.0+dfsg-1
4            r-cran-boolnet         2.1.3-1
5         r-cran-tikzdevice        0.10-1-1
6          r-cran-logspline         2.1.9-1
7            r-cran-genabel      1.8-0-1+b1
8                r-cran-lhs          0.14-1
9              r-bioc-limma   3.30.8+dfsg-1
10              r-cran-coin         1.1-3-1
11          r-other-iwrlars         0.9-5-2
12               r-cran-mnp         2.6-4-1
13               r-cran-msm         1.6.4-1
14            r-cran-fields          8.10-1
15           r-cran-desolve          1.14-1
16          r-cran-adephylo        1.1-10-2
17       r-cran-dosefinding        0.9-15-1
18            r-cran-deldir        0.1-12-1
19         r-cran-rniftilib    0.0-35.r79-2
20        r-cran-data.table        1.10.0-1
21               r-cran-qtl        1.40-8-1
22    r-bioc-preprocesscore        1.36.0-1
23          r-cran-contfrac        1.1-10-1
24            r-cran-glmnet         2.0-5-1
25            r-cran-bitops         1.0-6-1
26                r-cran-sp       1:1.2-4-1
27               r-cran-spc       1:0.5.3-1
28          r-bioc-snpstats   1.24.0+dfsg-1
29               r-cran-tgp        2.4-14-2
30             r-cran-brglm         0.5-9-1
31            r-cran-cmprsk         2.2-7-2
32              r-bioc-affy        1.52.0-1
33             r-cran-ncdf4       1.15-1+b2
34         r-cran-treescape       1.10.18-6
35              r-bioc-rbgl  1.50.0+dfsg1-1
36       r-bioc-rtracklayer        1.34.1-1
37            r-cran-hexbin        1.27.1-1
38         r-cran-princurve        1.1-12-1
39           r-cran-mapproj         1.2-4-1
40     r-cran-blockmodeling         0.1.8-1
41              r-cran-hdf5     1.6.10-4+b1
42              r-cran-pscl         1.4.9-1
43              r-cran-ade4         1.7-5-1
44              r-cran-vgam         1.0-3-1
45          r-cran-adegenet         2.0.1-1
46          r-cran-mixtools         1.0.4-1
47         r-cran-phylobase         0.8.2-1
48            r-cran-amelia         1.7.4-1
49              r-cran-spam         1.4-0-1
50      r-cran-medadherence          1.03-2
51           r-bioc-biobase        2.34.0-1
52      r-cran-surveillance        1.13.0-1
53 r-cran-randomfieldsutils        0.3.15-1
54              r-cran-deal      1:1.2-37-2
55        r-bioc-hilbertvis        1.32.0-1
56             r-cran-rcurl      1.95-4.8-2
57 r-other-mott-happy.hbrem           2.4-1
58          r-cran-mcmcpack         1.3-8-1
59          r-cran-spatstat        1.48-0-1
60             r-cran-vegan         2.4-2-1
61       r-cran-haplo.stats         1.7.7-1
62            r-cran-bayesm         3.0-2-2
63              r-cran-expm       0.999-0-1
64           r-bioc-dnacopy        1.48.0-1
65          r-cran-phangorn         2.1.1-1
66          r-cran-maptools 1:0.8-41+dfsg-1
67           r-cran-mlbench         2.1-1-1
68            r-bioc-affyio        1.44.0-1
69             r-bioc-graph        1.52.0-1
70           r-cran-polycub         0.5-2-1
71            r-bioc-deseq2        1.14.1-1
72          r-cran-pbivnorm         0.6.0-1
73             r-cran-caret  6.0-73+dfsg1-1
74        r-bioc-biovizbase        1.22.0-2
75              r-cran-nnls           1.4-1
76           r-cran-goftest         1.0-3-1
77            r-cran-igraph         1.0.1-1
78              r-cran-maps         3.1.1-1
79               r-cran-eco         3.1-7-1
80           r-cran-catools     1.17.1-1+b1
81      r-cran-randomfields        3.1.36-1
82               r-cran-erm        0.15-7-1
83               r-cran-etm         0.6-2-3
84      r-cran-randomforest        4.6-12-1
85               r-cran-evd         2.3-2-1
86      r-cran-raschsampler         0.8-8-1
87        r-bioc-genefilter        1.56.0-1
88              r-cran-mcmc         0.9-4-2
89             r-cran-spdep         0.6-9-1
90               r-cran-gam          1.14-1
91       r-other-amsmercury         1.3.0-2
92               r-cran-gsl      1.9-10.3-1
> 
```

With this, we can write out the content of the NMU request:

```{r, eval=FALSE}
> for (i in 1:nrow(res)) cat("nmu", paste(res[i,], collapse="_"), ". ANY . -m 'Rebuild against R 3.4.*, see #861333'\n")
nmu r-bioc-makecdfenv_1.50.0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-multtest_2.30.0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-edger_3.14.0+dfsg-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-boolnet_2.1.3-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-tikzdevice_0.10-1-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-logspline_2.1.9-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-genabel_1.8-0-1+b1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-lhs_0.14-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-limma_3.30.8+dfsg-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-coin_1.1-3-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-other-iwrlars_0.9-5-2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-mnp_2.6-4-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-msm_1.6.4-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-fields_8.10-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-desolve_1.14-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-adephylo_1.1-10-2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-dosefinding_0.9-15-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-deldir_0.1-12-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-rniftilib_0.0-35.r79-2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-data.table_1.10.0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-qtl_1.40-8-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-preprocesscore_1.36.0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-contfrac_1.1-10-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-glmnet_2.0-5-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-bitops_1.0-6-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-sp_1:1.2-4-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-spc_1:0.5.3-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-snpstats_1.24.0+dfsg-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-tgp_2.4-14-2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-brglm_0.5-9-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-cmprsk_2.2-7-2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-affy_1.52.0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-ncdf4_1.15-1+b2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-treescape_1.10.18-6 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-rbgl_1.50.0+dfsg1-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-rtracklayer_1.34.1-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-hexbin_1.27.1-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-princurve_1.1-12-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-mapproj_1.2-4-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-blockmodeling_0.1.8-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-hdf5_1.6.10-4+b1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-pscl_1.4.9-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-ade4_1.7-5-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-vgam_1.0-3-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-adegenet_2.0.1-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-mixtools_1.0.4-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-phylobase_0.8.2-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-amelia_1.7.4-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-spam_1.4-0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-medadherence_1.03-2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-biobase_2.34.0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-surveillance_1.13.0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-randomfieldsutils_0.3.15-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-deal_1:1.2-37-2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-hilbertvis_1.32.0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-rcurl_1.95-4.8-2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-other-mott-happy.hbrem_2.4-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-mcmcpack_1.3-8-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-spatstat_1.48-0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-vegan_2.4-2-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-haplo.stats_1.7.7-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-bayesm_3.0-2-2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-expm_0.999-0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-dnacopy_1.48.0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-phangorn_2.1.1-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-maptools_1:0.8-41+dfsg-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-mlbench_2.1-1-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-affyio_1.44.0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-graph_1.52.0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-polycub_0.5-2-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-deseq2_1.14.1-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-pbivnorm_0.6.0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-caret_6.0-73+dfsg1-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-biovizbase_1.22.0-2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-nnls_1.4-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-goftest_1.0-3-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-igraph_1.0.1-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-maps_3.1.1-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-eco_3.1-7-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-catools_1.17.1-1+b1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-randomfields_3.1.36-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-erm_0.15-7-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-etm_0.6-2-3 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-randomforest_4.6-12-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-evd_2.3-2-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-raschsampler_0.8-8-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-genefilter_1.56.0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-mcmc_0.9-4-2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-spdep_0.6-9-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-gam_1.14-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-other-amsmercury_1.3.0-2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-gsl_1.9-10.3-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
> 
```

## Summary

The finalu set of 92 NMUs is the minimal change required, and reasonable relative to the over 
500 reverse dependencies of R itself.  We are able to narrow the set of packages requiring
a rebuild down by a combining data from the R package system, the Debian package system
and (some) package sources we were able to access on a CRAN-related server.
