---
title: "Minimal Set of binnmu Packages"
subtitle: "Combining the R and Debian package systems"
author: "Dirk Eddelbuettel"
date: "First version 2017-Jul-16; this version 2017-Aug-05"
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
vignette: >
  %\VignetteIndexEntry{Minimal Set of binnmu Packages}
  %\VignettePackage{RcppAPT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Step 0: Problem Definition

### Upstream Change

R 3.4.0, released in April, included the following paragraph in its NEWS file:

 * Packages which register native routines for .C or .Fortran need
   to be re-installed for this version (unless installed with
   R-devel SVN revision r72375 or later).

This transition has no fallback behavior (as is more common with R changes) 
and requires a rebuild.  Packages build under older R version still _load_ 
and function partially, but will be unable to access any native (_i.e._, compiled)
routines.

### Impact

For the Debian packages, this means that we need to consider the set of packages which

- match `r-cran-*`, `r-bioc-*` and alike
- contain compiled code (as R-only packages have no native routines)
- use at least one `.C()` or `.Fortran()` (but _not_ `.Call()`) call
- use (the hitherto optional) routine registration (so that the change in behaviour is noticible)
- have _not yet_ been recompiled with R 3.4.0 or R 3.4.1

This note computes this set and provides the input for a 
[wanna-build](https://release.debian.org/wanna-build.txt) request.

This version is updated version which reflects the fourth point above which was pointed out to
me by Kurt Hornik after I shared the initial version with him.  The point he raised ("does it use 
`R_registerRoutines` ?") is important and further reduces the effective set.

## Step 1: Reverse Dependencies of R

### Fresh Debian unstable session

For this we drop into a clean Docker container running Debian unstable. Later, we will need
the current sources of the [`RcppAPT`](https://github.com/eddelbuettel/rcppapt) package so 
we start from a local git directory:

```sh
server> cd ~/git && docker run --rm -ti -v $(pwd):/mnt debian:unstable
```

### Update Debian

Inside the Docker container, we update the package information and install what
is needed to build [`RcppAPT`](https://github.com/eddelbuettel/rcppapt) for R.  
This includes Rcpp and `libapt-pkg-dev`.  We also install the data.table package
used for aggregating the (R and Debian) package data computed below.

This step takes a short moment, with the exact time dependent on the network 
connection and other factors.

```sh
docker> apt-get update
docker> apt-get -y dist-upgrade

docker> apt-get -y install r-cran-rcpp r-cran-data.table libapt-pkg-dev less 

docker> cd /mnt 
docker> R CMD INSTALL rcppapt/       # assuming we're above rcppapt
```

### Launch R

#### All Candidates

Inside the same Docker session, we now launch R and run (almost all of) the remainder from R.

```{r, eval=FALSE}
> library(RcppAPT)
> library(data.table)
> rd <- reverseDepends("r-base-core")         # 516 x 2
> rd <- rd[grepl("^r-", rd[,1]), ]            # 489 x 2
> rd <- rd[order(rd[,2]), ]
> setDT(rd)
```

We use `RcppAPT` to compute the reverse depends of the main R package providing the R engine: `r-base-core`.
Among those (currently) 516 packages are both other packages from the upstream source (`r-base*`, `r-doc*`) which
we exclude first as well as other, non-R-package dependencies (such as `rpy2`) which we also exclude.

This leaves 489 candidate packages out of the initial 514.  The version field tells which r-base-core version
was used to build the package---information we need per the setup described above.

```{r, eval=FALSE}
> rd
> rd
             package version
  1:       r-doc-pdf
  2:      r-doc-info
  3:      r-doc-html
  4:     r-base-html
  5:     r-other-rot
 ---
485: r-base-core-dbg 3.4.1-2
486:          r-base 3.4.1-2
487:     r-cran-mgcv 3.4.1-2
488:     r-cran-boot 3.4.1-2
489:      r-cran-car 3.4.1-2
> 
```

Next we need to filter out two versions with unsortable (_i.e._,non-semantic) version numbers,
and apply a logical filter depending on whether the package was built with R version 3.3.3 or 
earlier, indicating a possibe required rebuild.

```{r, eval=FALSE}
> rd[ version=="3.0.0~20130330-1", version := "3.0.0.20130330-1"]
> rd[ version=="3.2.4-revised-1", version := "3.2.4.1-1"]
> rd[version!="", oldVersion := version  <=  package_version("3.3.3-1")]
> rd[ is.na(oldVersion), oldVersion := FALSE]
> rd[ !grepl("r-(doc|base)", package), ]
             package          version oldVersion
  1:     r-other-rot                       FALSE
  2: r-cran-epitools          3.0.0-2       TRUE
  3: r-cran-combinat          3.0.0-2       TRUE
  4:    r-cran-gmaps 3.0.0.20130330-1       TRUE
  5:      r-cran-wdi          3.0.1-6       TRUE
 ---
478:   r-recommended          3.4.1-2      FALSE
479:       r-mathlib          3.4.1-2      FALSE
480:     r-cran-mgcv          3.4.1-2      FALSE
481:     r-cran-boot          3.4.1-2      FALSE
482:      r-cran-car          3.4.1-2      FALSE
```

To cover some corner case, we derive a `skip` field:

```{r, eval=FALSE}
> rd[ version=="", skip:=TRUE ]
> rd[ is.na(skip), skip:=FALSE]
> rd[ skip==FALSE, ]
             package          version oldVersion  skip
  1: r-cran-epitools          3.0.0-2       TRUE FALSE
  2: r-cran-combinat          3.0.0-2       TRUE FALSE
  3:    r-cran-gmaps 3.0.0.20130330-1       TRUE FALSE
  4:      r-cran-wdi          3.0.1-6       TRUE FALSE
  5:   r-cran-bitops          3.0.1-6       TRUE FALSE
 ---
480: r-base-core-dbg          3.4.1-2      FALSE FALSE
481:          r-base          3.4.1-2      FALSE FALSE
482:     r-cran-mgcv          3.4.1-2      FALSE FALSE
483:     r-cran-boot          3.4.1-2      FALSE FALSE
484:      r-cran-car          3.4.1-2      FALSE FALSE
> 
```

#### Compiled Packages


Next, we find the actual dependencies of each of these packages by constructing
a large regular expression which we feed into `RcppAPT::getDepends()`


```{r, eval=FALSE}
> regexp <- paste(paste0("^", rd[skip==FALSE, package], "$"), collapse="|")
> dep <- getDepends(regexp)
> setDT(dep)
> dep
                  srcpkg              deppkg cmpop          version
   1:  r-bioc-hypergraph         r-base-core     2 3.3.1.20161024-1
   2:  r-bioc-hypergraph             r-api-3     0           (null)
   3:  r-bioc-hypergraph        r-bioc-graph     0           (null)
   4:  r-bioc-hypergraph r-bioc-biocgenerics     0           (null)
   5:  r-bioc-hypergraph        r-cran-runit     0           (null)
  ---
3744: r-cran-viridislite             r-api-3     0           (null)
3745:      r-cran-xtable         r-base-core     2          3.2.5-1
3746:      r-cran-xtable             r-api-3     0           (null)
3747:   r-cran-pkgkitten         r-base-core     2          3.3.2-1
3748:   r-cran-pkgkitten             r-api-3     0           (null)
> 
```

Next we subset to those have `libc6` as a Depends, meaning they are compiled packages.
This excludes all the R packages having only R code.

```{r, eval=FALSE}
> comp <- dep[deppkg=="libc6"]   # 242
> comp
                 srcpkg deppkg cmpop version isCompiled
  1:  r-bioc-makecdfenv  libc6     2     2.4       TRUE
  2:       r-cran-bio3d  libc6     2    2.14       TRUE
  3:   r-bioc-rsamtools  libc6     2    2.15       TRUE
  4:     r-cran-foreign  libc6     2    2.14       TRUE
  5:    r-bioc-multtest  libc6     2    2.14       TRUE
 ---
238:     r-cran-nleqslv  libc6     2     2.4       TRUE
239: r-other-amsmercury  libc6     2    2.14       TRUE
240:         r-cran-gnm  libc6     2     2.4       TRUE
241:         r-cran-gsl  libc6     2     2.4       TRUE
242:         r-cran-gss  libc6     2     2.4       TRUE
>
```

We are now getting closer.  We set keys on the `data.table` objects, and then do 
an inner join:

```{r, eval=FALSE}
> setkey(comp, srcpkg)
> setkey(rd, package)
> all <- rd[comp[, c(1,5)]]   # inner join (by default on columns with keys)
> all[order(version),]
                      package version oldVersion  skip isCompiled
  1:            r-cran-bitops 3.0.1-6       TRUE FALSE       TRUE
  2:               r-cran-mnp 3.0.2-1       TRUE FALSE       TRUE
  3: r-other-mott-happy.hbrem 3.0.2-1       TRUE FALSE       TRUE
  4:             r-cran-amore 3.1.0-1       TRUE FALSE       TRUE
  5:              r-cran-deal 3.1.0-1       TRUE FALSE       TRUE
 ---
238:              r-cran-rcpp 3.4.1-2      FALSE FALSE       TRUE
239:            r-cran-rmysql 3.4.1-2      FALSE FALSE       TRUE
240:         r-cran-rsymphony 3.4.1-2      FALSE FALSE       TRUE
241:               r-cran-ttr 3.4.1-2      FALSE FALSE       TRUE
242:                r-mathlib 3.4.1-2      FALSE FALSE       TRUE
> 
```

We have 242 _potential_ rebuilds, down from 514 reverse depends at the outset.

#### Version check

Next, we can concentrate on those having been built with the older versions requiring
a rebuild:


```{r, eval=FALSE}
> all[oldVersion==TRUE,][order(version),]    # 167
                      package version oldVersion  skip isCompiled
  1:            r-cran-bitops 3.0.1-6       TRUE FALSE       TRUE
  2:               r-cran-mnp 3.0.2-1       TRUE FALSE       TRUE
  3: r-other-mott-happy.hbrem 3.0.2-1       TRUE FALSE       TRUE
  4:             r-cran-amore 3.1.0-1       TRUE FALSE       TRUE
  5:              r-cran-deal 3.1.0-1       TRUE FALSE       TRUE
 ---                                                             
163:           r-cran-rcppgsl 3.3.3-1       TRUE FALSE       TRUE
164:             r-cran-rodbc 3.3.3-1       TRUE FALSE       TRUE
165:         r-cran-snowballc 3.3.3-1       TRUE FALSE       TRUE
166:                r-cran-v8 3.3.3-1       TRUE FALSE       TRUE
167:               r-cran-zoo 3.3.3-1       TRUE FALSE       TRUE
> 
```

Now we are down to 167 packages.

```{r, eval=FALSE}
> all[, cran:=grepl("^r-cran", package) ]
> all[, bioc:=grepl("^r-bioc", package) ]
> all[bioc==TRUE & oldVersion==TRUE,]                # 17 BioC
                  package          version oldVersion  skip isCompiled  cran bioc
 1:           r-bioc-affy 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
 2:         r-bioc-affyio 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
 3:        r-bioc-biobase 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
 4:     r-bioc-biovizbase          3.3.2-1       TRUE FALSE       TRUE FALSE TRUE
 5:         r-bioc-deseq2          3.3.2-1       TRUE FALSE       TRUE FALSE TRUE
 6:        r-bioc-dnacopy 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
 7:          r-bioc-edger          3.3.0-2       TRUE FALSE       TRUE FALSE TRUE
 8:     r-bioc-genefilter          3.3.2-1       TRUE FALSE       TRUE FALSE TRUE
 9:          r-bioc-graph 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
10:     r-bioc-hilbertvis 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
11:          r-bioc-limma          3.3.2-1       TRUE FALSE       TRUE FALSE TRUE
12:     r-bioc-makecdfenv 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
13:       r-bioc-multtest 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
14: r-bioc-preprocesscore 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
15:           r-bioc-rbgl          3.3.2-1       TRUE FALSE       TRUE FALSE TRUE
16:    r-bioc-rtracklayer          3.3.2-1       TRUE FALSE       TRUE FALSE TRUE
17:       r-bioc-snpstats 3.3.1.20161024-1       TRUE FALSE       TRUE FALSE TRUE
> 
```

Among these are 17 BioConductor packages.  This is a superset as we do not know which of these 
use only `.Call()` meaning that no rebuild would be required.   


```{r, eval=FALSE}
> all[bioc!=TRUE & cran!=TRUE & oldVersion==TRUE,]   # 3 other
                    package version oldVersion  skip isCompiled  cran  bioc
1:       r-other-amsmercury 3.3.2-1       TRUE FALSE       TRUE FALSE FALSE
2:          r-other-iwrlars 3.3.2-1       TRUE FALSE       TRUE FALSE FALSE
3: r-other-mott-happy.hbrem 3.0.2-1       TRUE FALSE       TRUE FALSE FALSE
> 
```

There are also three which are neither BioC nor CRAN.

```{r, eval=FALSE}
> cand <- all[ cran==TRUE & oldVersion==TRUE, ]   # 147
> cand
             package version oldVersion  skip isCompiled cran  bioc
  1:     r-cran-ade4 3.3.2-1       TRUE FALSE       TRUE TRUE FALSE
  2: r-cran-adegenet 3.3.1-1       TRUE FALSE       TRUE TRUE FALSE
  3: r-cran-adephylo 3.3.2-1       TRUE FALSE       TRUE TRUE FALSE
  4:   r-cran-amelia 3.2.3-1       TRUE FALSE       TRUE TRUE FALSE
  5:    r-cran-amore 3.1.0-1       TRUE FALSE       TRUE TRUE FALSE
 ---                                                               
143:    r-cran-vegan 3.3.2-1       TRUE FALSE       TRUE TRUE FALSE
144:     r-cran-vgam 3.3.2-1       TRUE FALSE       TRUE TRUE FALSE
145:     r-cran-xml2 3.3.2-1       TRUE FALSE       TRUE TRUE FALSE
146:     r-cran-yaml 3.3.2-1       TRUE FALSE       TRUE TRUE FALSE
147:      r-cran-zoo 3.3.3-1       TRUE FALSE       TRUE TRUE FALSE
> 
```

We have 147 possible NMUs based off CRAN.

Next, we mix this with information from CRAN.

```{r, eval=FALSE}
> db <- tools::CRAN_package_db()   # CRAN pkge info: N rows x 65 cols
> setDT(db)
> db[, package:=paste0("r-cran-", tolower(Package))]
> setkey(db, package)              # key on package field
> foo <- db[ cand ]                # inner join
> foo[, .(package, Package, Version, NeedsCompilation, oldVersion, skip)]
             package  Package Version NeedsCompilation oldVersion  skip
  1:     r-cran-ade4     ade4   1.7-6              yes       TRUE FALSE
  2: r-cran-adegenet adegenet   2.0.1              yes       TRUE FALSE
  3: r-cran-adephylo adephylo  1.1-10              yes       TRUE FALSE
  4:   r-cran-amelia   Amelia   1.7.4              yes       TRUE FALSE
  5:    r-cran-amore    AMORE  0.2-15              yes       TRUE FALSE
 ---                                                                   
143:    r-cran-vegan    vegan   2.4-3              yes       TRUE FALSE
144:     r-cran-vgam     VGAM   1.0-3              yes       TRUE FALSE
145:     r-cran-xml2     xml2   1.1.1              yes       TRUE FALSE
146:     r-cran-yaml     yaml  2.1.14              yes       TRUE FALSE
147:      r-cran-zoo      zoo   1.8-0              yes       TRUE FALSE
> 
```

This is our set of 147 candidate packages with their CRAN name, Debian name and upstream version.

```{r, eval=FALSE}
> saveRDS(foo[, .(package, Package, Version, NeedsCompilation, oldVersion, skip)], file="debpackages.rds")
```

We save this file to be used on another machine.

## Step 2: Grep

On another machine with access to all CRAN package sources (which I happen to have access to), we use the list 
of 147 candidate packages and run a recursive grep for each.  We store the output from two `egrep` runs, called via `system()`, 
directly in the same data structure.  The first checks for `.C()` or `.Fortran()` calls in the R scripts; the second checks
for `R_registerRoutines()` in the compiled C code (with thanks again to Kurt Hornik for the suggestion)

```{r, eval=FALSE}
deb <- readRDS("~/debpackages.rds")
for (i in 1:nrow(deb)) {
    deb[i, "dotCorFortran"] <- if (is.na(deb[i, "Package"])) NA 
                               else system(paste0("egrep -r -q \"\\.(C|Fortran)\\(\" ", deb[i, "Package"], "/R/*"))==0
    deb[i, "hasRegistration"] <- if (is.na(deb[i, "Package"])) NA 
                               else system(paste0("egrep -r -q \"R_registerRoutines\\(\" ", deb[i, "Package"], "/src/*"))==0
}
saveRDS(deb, "~/debpackagesout.rds")
```

## Step 3: Finalize

We read the data back in and subset on those for which the recursive grep found actual uses of
`.C()` or `.Fortran()`.  The list contains 72 packages.

```{r, eval=FALSE}
> deb <- readRDS("debpackagesout.rds")
> setDT(deb)
> deb[ is.na(dotCorFortran) |(dotCorFortran & hasRegistration), 1:3]                               
                     package           Package  Version
 1:              r-cran-ade4              ade4    1.7-6
 2:            r-cran-bayesm            bayesm  3.1-0.1
 3:     r-cran-blockmodeling     blockmodeling    0.1.9
 4:             r-cran-brglm             brglm    0.6.1
 5:             r-cran-caret             caret   6.0-76
 6:              r-cran-coin              coin    1.2-1
 7:          r-cran-contfrac          contfrac   1.1-11
 8:        r-cran-data.table        data.table   1.10.4
 9:            r-cran-deldir            deldir   0.1-14
10:           r-cran-desolve           deSolve     1.20
11:               r-cran-eco               eco    4.0-1
12:              r-cran-expm              expm  0.999-2
13:            r-cran-fields            fields      9.0
14:               r-cran-gam               gam   1.14-4
15:            r-cran-glmnet            glmnet   2.0-10
16:           r-cran-goftest           goftest    1.1-1
17:              r-cran-hdf5                NA       NA
18:            r-cran-igraph            igraph    1.1.2
19:           r-cran-mapproj           mapproj    1.2-5
20:              r-cran-maps              maps    3.2.0
21:          r-cran-maptools          maptools    0.9-2
22:              r-cran-mcmc              mcmc    0.9-5
23:          r-cran-mcmcpack          MCMCpack    1.4-0
24:      r-cran-medadherence                NA       NA
25:          r-cran-mixtools          mixtools    1.1.0
26:               r-cran-mnp               MNP    3.0-2
27:             r-cran-ncdf4             ncdf4     1.16
28:          r-cran-phangorn          phangorn    2.2.0
29:         r-cran-phylobase         phylobase    0.8.4
30:               r-cran-qtl               qtl   1.41-6
31:      r-cran-randomfields      RandomFields   3.1.50
32: r-cran-randomfieldsutils RandomFieldsUtils   0.3.25
33:             r-cran-rcurl             RCurl 1.95-4.8
34:         r-cran-rniftilib                NA       NA
35:                r-cran-sp                sp    1.2-5
36:              r-cran-spam              spam    2.1-1
37:          r-cran-spatstat          spatstat   1.51-0
38:             r-cran-spdep             spdep   0.6-13
39:      r-cran-surveillance      surveillance   1.14.0
40:         r-cran-treescape                NA       NA
41:             r-cran-vegan             vegan    2.4-3
42:              r-cran-vgam              VGAM    1.0-4
                     package           Package  Version
> 
```

Similarly, the 17 BioC and 3 other packages can be tested via recursive greps (not shown) in a directory
filled with `apt-get source` downloads:

```{r, eval=FALSE}
pkgs <- rbind(all[bioc!=TRUE & cran!=TRUE & oldVersion==TRUE, 1],
              all[bioc==TRUE & oldVersion==TRUE, 1])[[1]]

dir.create("/tmp/scratch")
setwd("/tmp/scratch")

cat("deb-src http://deb.debian.org/debian unstable main\n",
    file="/etc/apt/sources.list", append=TRUE)
system("apt-get update")

for (p in pkgs) system(paste("apt-get source", p))

df <- data.frame(package=pkgs, stringsAsFactors=FALSE)
for (i in 1:nrow(df)) {
    p <- df[i, 1]
    df[i, "dotCorFortran"] <- system(paste0("egrep -r -q \"\\.(C|Fortran)\\(\" ", p, "*/R/*"))==0
    df[i, "hasRegistration"] <- system(paste0("egrep -r -q \"R_registerRoutines\\(\" ", p, "*/src/*"))==0
}

setDT(df)
```

This leads to a further four packages:

```{r, eval=FALSE}
> df[dotCorFortran & hasRegistration, 1]     
                     pkg
1:           r-bioc-affy
2:          r-bioc-edger
3:     r-bioc-genefilter
4: r-bioc-preprocesscore
> 
```

These 42, along with the 4 (from the initally 17 BioC and 3 'other') packages are our target set.

```{r, eval=FALSE}
> nmu <- deb[ is.na(dotCorFortran) | (dotCorFortran & hasRegistration), 1] #42
> oth <- df[dotCorFortran & hasRegistration, 1]
>
> nmu <- rbind(nmu, oth)  ## 46
> nmu
                     package
 1:              r-cran-ade4
 2:            r-cran-bayesm
 3:     r-cran-blockmodeling
 4:             r-cran-brglm
 5:             r-cran-caret
 6:              r-cran-coin
 7:          r-cran-contfrac
 8:        r-cran-data.table
 9:            r-cran-deldir
10:           r-cran-desolve
11:               r-cran-eco
12:              r-cran-expm
13:            r-cran-fields
14:               r-cran-gam
15:            r-cran-glmnet
16:           r-cran-goftest
17:              r-cran-hdf5
18:            r-cran-igraph
19:           r-cran-mapproj
20:              r-cran-maps
21:          r-cran-maptools
22:              r-cran-mcmc
23:          r-cran-mcmcpack
24:      r-cran-medadherence
25:          r-cran-mixtools
26:               r-cran-mnp
27:             r-cran-ncdf4
28:          r-cran-phangorn
29:         r-cran-phylobase
30:               r-cran-qtl
31:      r-cran-randomfields
32: r-cran-randomfieldsutils
33:             r-cran-rcurl
34:         r-cran-rniftilib
35:                r-cran-sp
36:              r-cran-spam
37:          r-cran-spatstat
38:             r-cran-spdep
39:      r-cran-surveillance
40:         r-cran-treescape
41:             r-cran-vegan
42:              r-cran-vgam
43:              r-bioc-affy
44:             r-bioc-edger
45:        r-bioc-genefilter
46:    r-bioc-preprocesscore
                     package
>
> 
```

We need to retrieve the version number in Debian unstable of these packages by once agaim
relying of a function from [`RcppAPT`](https://github.com/eddelbuettel/rcppapt) 

```{r, eval=FALSE}
> regexp <- paste(paste0("^", nmu[[1]], "$"), collapse="|")
>
> res <- getPackages(regexp)
> res
                    Package         Version
1              r-bioc-edger   3.14.0+dfsg-1
2               r-cran-coin         1.1-3-1
3                r-cran-mnp         2.6-4-1
4             r-cran-fields          8.10-1
5            r-cran-desolve          1.14-1
6             r-cran-deldir        0.1-12-1
7          r-cran-rniftilib    0.0-35.r79-2
8         r-cran-data.table        1.10.0-1
9                r-cran-qtl        1.40-8-1
10    r-bioc-preprocesscore        1.36.0-1
11          r-cran-contfrac        1.1-10-1
12            r-cran-glmnet         2.0-5-1
13                r-cran-sp       1:1.2-4-1
14             r-cran-brglm         0.5-9-1
15              r-bioc-affy        1.52.0-1
16             r-cran-ncdf4       1.15-1+b2
17         r-cran-treescape       1.10.18-6
18           r-cran-mapproj         1.2-4-1
19     r-cran-blockmodeling         0.1.8-1
20              r-cran-hdf5     1.6.10-4+b1
21              r-cran-ade4         1.7-5-1
22              r-cran-vgam         1.0-3-1
23          r-cran-mixtools         1.0.4-1
24         r-cran-phylobase         0.8.2-1
25              r-cran-spam         1.4-0-1
26      r-cran-medadherence          1.03-2
27      r-cran-surveillance        1.13.0-1
28 r-cran-randomfieldsutils        0.3.15-1
29             r-cran-rcurl      1.95-4.8-2
30          r-cran-mcmcpack         1.3-8-1
31          r-cran-spatstat        1.48-0-1
32             r-cran-vegan         2.4-2-1
33            r-cran-bayesm         3.0-2-2
34              r-cran-expm       0.999-0-1
35          r-cran-phangorn         2.1.1-1
36          r-cran-maptools 1:0.8-41+dfsg-1
37             r-cran-caret  6.0-73+dfsg1-1
38           r-cran-goftest         1.0-3-1
39            r-cran-igraph         1.0.1-1
40              r-cran-maps         3.1.1-1
41               r-cran-eco         3.1-7-1
42      r-cran-randomfields        3.1.36-1
43        r-bioc-genefilter        1.56.0-1
44              r-cran-mcmc         0.9-4-2
45             r-cran-spdep         0.6-9-1
46               r-cran-gam          1.14-1
>
```

With this, we can write out the content of the NMU request:

```{r, eval=FALSE}
> 
> for (i in 1:nrow(res))
+     cat("nmu", paste(res[i,], collapse="_"), ". ANY . -m 'Rebuild against R 3.4.*, see #861333'\n")
nmu r-bioc-edger_3.14.0+dfsg-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-coin_1.1-3-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-mnp_2.6-4-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-fields_8.10-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-desolve_1.14-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-deldir_0.1-12-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-rniftilib_0.0-35.r79-2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-data.table_1.10.0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-qtl_1.40-8-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-preprocesscore_1.36.0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-contfrac_1.1-10-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-glmnet_2.0-5-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-sp_1:1.2-4-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-brglm_0.5-9-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-affy_1.52.0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-ncdf4_1.15-1+b2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-treescape_1.10.18-6 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-mapproj_1.2-4-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-blockmodeling_0.1.8-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-hdf5_1.6.10-4+b1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-ade4_1.7-5-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-vgam_1.0-3-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-mixtools_1.0.4-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-phylobase_0.8.2-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-spam_1.4-0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-medadherence_1.03-2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-surveillance_1.13.0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-randomfieldsutils_0.3.15-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-rcurl_1.95-4.8-2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-mcmcpack_1.3-8-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-spatstat_1.48-0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-vegan_2.4-2-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-bayesm_3.0-2-2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-expm_0.999-0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-phangorn_2.1.1-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-maptools_1:0.8-41+dfsg-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-caret_6.0-73+dfsg1-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-goftest_1.0-3-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-igraph_1.0.1-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-maps_3.1.1-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-eco_3.1-7-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-randomfields_3.1.36-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-bioc-genefilter_1.56.0-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-mcmc_0.9-4-2 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-spdep_0.6-9-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
nmu r-cran-gam_1.14-1 . ANY . -m 'Rebuild against R 3.4.*, see #861333'
>
```

## Summary

The final set of 46 NMUs is the minimal change required, and reasonable relative to the 
516 reverse dependencies of R itself.  We are able to narrow the set of packages requiring
a rebuild down by a combining data from the R package system, the Debian package system
and (some) package sources we were able to access on a CRAN-related server.

### Acknowledgements 

Thanks for Kurt Hornik for pointing out the additional check for `R_registerRoutine` in the in C code, leading
to a further reduction from 90+ packages to 46.

### History

The first published version (Julyu 2017) did not check for `R_registerRoutines`. The second version (August 2017) 
does, leading to 46 suggested NMUs.

### See Also

The [source file](https://github.com/eddelbuettel/rcppapt/blob/master/vignettes/binnmuAfterR340.Rmd) is on GitHub 
as is the  [revision history](https://github.com/eddelbuettel/rcppapt/commits/master/vignettes/binnmuAfterR340.Rmd).
The [corresponding Debian bug report](https://bugs.debian.org/868558) is based on this analysis.
